---
title: "E6690 HW1"
# header-includes:
#    - \setlength\parindent{20pt}
output:
  pdf_document:
    latex_engine: xelatex
---

This file contains the how we train the GBM model for the prediction of human behavior. 
We use h2o package which can speedup the training process by exploiting the parallelism in PC.
The brief procedure in training this model: 
  load the datasets;
  do the data preprocessing;
  demension reduction using PCA and do some data visualization;
  train the GBM model using the feature extracted data;
  test the model to see the accuracy.


At the end of the file, we also attached how to do the grid search to find the best parameters to train the model in HAR problem setting.



```{r}
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(h2o))
suppressPackageStartupMessages(library(plotly))

# Start and connect to a H2O cluster (JVM)
h2o.init(nthreads = -1)
```



```{r}
# Import datasets
hex_train <- h2o.importFile("./dataset/train.csv.gz")
hex_test <- h2o.importFile("./dataset/test.csv.gz")
```




```{r}
# Extract 'activity' columns for other graphics packages in R
d_activity_train <- as.data.frame(hex_train$activity)
d_activity_test <- as.data.frame(hex_test$activity)
# Count acitivity 
d_freq_train <- as.data.frame(table(d_activity_train))
d_freq_test <- as.data.frame(table(d_activity_test))
d_freq <- merge(d_freq_train, d_freq_test, by.x = "d_activity_train", by.y = "d_activity_test", sort = FALSE)
colnames(d_freq) <- c("activity", "freq_train", "freq_test")
d_freq

```



```{r}
# Principal Component Analysis
# 95% of variance in original data captured by first five principal components
model_pca <- h2o.prcomp(training_frame = hex_train, 
                    x = 2:562, 
                    model_id = "h2o_pca",
                    k = 5)
model_pca    
```




```{r}
# Visualize principle components with activity labels
d_pca <- as.data.frame(h2o.predict(model_pca, hex_train))
d_pca <- data.frame(d_pca, as.data.frame(hex_train$activity))
p <- plot_ly(data = d_pca, x = ~PC1, y = ~PC2, z = ~PC3, color = ~activity) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'PC1'),
                     yaxis = list(title = 'PC2'),
                     zaxis = list(title = 'PC3')))
p
```




```{r}

p <- plot_ly(data = d_pca, x = ~PC1, y = ~PC2, color = ~activity, 
             type = "scatter", mode = "markers", marker = list(size = 3)) %>%
     layout(title = "Visualizing Principle Components")
p
# export(p, 'HAR_pca_pc2&3.png')
```



Step 4 - Build and evalutate a predictive model using H2O’s Gradient Boosting Machine (GBM) algorithm

```{r}
# Define target and features for model training
target <- "activity"
features <- setdiff(colnames(hex_train), target) # i.e. using the records of all 561 sensors
```

```{r}
# Build a GBM model with cross-validation and early stopping
model <- h2o.gbm(x = features,
                 y = target,
                 training_frame = hex_train,                 
                 model_id = "h2o_gbm",
                 ntrees = 500,
                 learn_rate = 0.05,
                 learn_rate_annealing = 0.999,
                 max_depth = 22,
                 sample_rate = 0.64,
                 col_sample_rate = 0.5,
                 nfolds = 7,
                 fold_assignment = "Stratified",
                 histogram_type = "RoundRobin",
                 stopping_metric = "logloss",
                 stopping_rounds = 5,
                 score_tree_interval = 10,
                 seed = 1234)

# Print out model summary
model
```


```{r}
# Visualize variable importance
h2o.varimp_plot(model, num_of_features = 15)
```

```{r}
# Make predictions
yhat_test <- h2o.predict(model, hex_test)
head(yhat_test)
```

```{r}
# Evaluate predictions
h2o.performance <- h2o.performance(model, newdata = hex_test)
h2o.performance
```



## Search for hyper-parameter

```{r}
# search for maxdepth
hyper_params = list( max_depth = seq(1,29,2) )
grid <- h2o.grid(
  ## hyper parameters
  hyper_params = hyper_params,
  
  ## full Cartesian hyper-parameter search
  search_criteria = list(strategy = "Cartesian"),
  
  ## which algorithm to run
  algorithm="gbm",
  
  ## identifier for the grid, to later retrieve it
  grid_id="depth_grid",
  
  ## standard model parameters
  x = features, 
  y = target, 
  training_frame = hex_train[1:1000,], 
  
  ## more trees is better if the learning rate is small enough 
  ## here, use "more than enough" trees - we have early stopping
  ntrees = 1000,                                                            
  
  ## smaller learning rate is better
  ## since we have learning_rate_annealing, we can afford to start with a bigger learning rate
  learn_rate = 0.05,                                                         
  
  ## learning rate annealing: learning_rate shrinks by 1% after every tree 
  ## (use 1.00 to disable, but then lower the learning_rate)
  learn_rate_annealing = 0.99,                                               
  
  ## sample 80% of rows per tree
  sample_rate = 0.8,                                                       

  ## sample 80% of columns per split
  col_sample_rate = 0.8, 
## fix a random number generator seed for reproducibility
  seed = 1234,                                                             
  
  ## early stopping once the validation AUC doesn't improve by at least 0.01% for 5 consecutive scoring events
  stopping_rounds = 5,
  stopping_tolerance = 1e-4,
  stopping_metric = "logloss", 
  
  ## score every 10 trees to make early stopping reproducible (it depends on the scoring interval)
  score_tree_interval = 10                                                
)

## by default, display the grid search results sorted by increasing logloss (since this is a classification task)
grid                                                                       

## sort the grid models by decreasing logloss
sortedGrid <- h2o.getGrid("depth_grid", sort_by="logloss", decreasing = TRUE)    
sortedGrid

## find the range of max_depth for the top 5 models
topDepths = sortedGrid@summary_table$max_depth[1:5]                       
minDepth = min(as.numeric(topDepths))
maxDepth = max(as.numeric(topDepths))
minDepth
maxDepth






```


Now that we know a good range for max_depth, we can tune all other parameters in more detail. Since we don't know what combinations of hyper-parameters will result in the best model, we'll use random hyper-parameter search to "let the machine get luckier than a best guess of any human".


```{r}
hyper_params = list( 
  ## restrict the search to the range of max_depth established above
  max_depth = seq(minDepth,maxDepth,1),                                      
  
  ## search a large space of row sampling rates per tree
  sample_rate = seq(0.2,1,0.01),                                             
  
  ## search a large space of column sampling rates per split
  col_sample_rate = seq(0.2,1,0.01),                                         
  
  ## search a large space of column sampling rates per tree
  col_sample_rate_per_tree = seq(0.2,1,0.01),                                
  
  ## search a large space of how column sampling per split should change as a function of the depth of the split
  col_sample_rate_change_per_level = seq(0.9,1.1,0.01),                      
  
  ## search a large space of the number of min rows in a terminal node
  min_rows = 2^seq(0,log2(nrow(hex_train[1:1000,]))-1,1),                                 
  
  ## search a large space of the number of bins for split-finding for continuous and integer columns
  nbins = 2^seq(4,10,1),                                                     
  
  ## search a large space of the number of bins for split-finding for categorical columns
  nbins_cats = 2^seq(4,12,1),                                                
  
  ## search a few minimum required relative error improvement thresholds for a split to happen
  min_split_improvement = c(0,1e-8,1e-6,1e-4),                               
  
  ## try all histogram types (QuantilesGlobal and RoundRobin are good for numeric columns with outliers)
  histogram_type = c("UniformAdaptive","QuantilesGlobal","RoundRobin")       
)

search_criteria = list(
  ## Random grid search
  strategy = "RandomDiscrete",      
  
  ## limit the runtime to 60 minutes
  max_runtime_secs = 3600,         
  
  ## build no more than 100 models
  max_models = 100,                  
  
  ## random number generator seed to make sampling of parameter combinations reproducible
  seed = 1234,                        
  
  ## early stopping once the leaderboard of the top 5 models is converged to 0.1% relative difference
  stopping_rounds = 5,                
  stopping_metric = "logloss",
  stopping_tolerance = 1e-3
)

grid <- h2o.grid(
  ## hyper parameters
  hyper_params = hyper_params,
  
  ## hyper-parameter search configuration (see above)
  search_criteria = search_criteria,
  
  ## which algorithm to run
  algorithm = "gbm",
  
  ## identifier for the grid, to later retrieve it
  grid_id = "final_grid", 
  
  ## standard model parameters
  x = features, 
  y = target, 
  training_frame = hex_train[1:1000,], 
  
  ## more trees is better if the learning rate is small enough
  ## use "more than enough" trees - we have early stopping
  ntrees = 1000,                                                            
  
  ## smaller learning rate is better
  ## since we have learning_rate_annealing, we can afford to start with a bigger learning rate
  learn_rate = 0.05,                                                         
  
  ## learning rate annealing: learning_rate shrinks by 1% after every tree 
  ## (use 1.00 to disable, but then lower the learning_rate)
  learn_rate_annealing = 0.99,                                               
  
  ## early stopping based on timeout (no model should take more than 1 hour - modify as needed)
  max_runtime_secs = 3600,                                                 
  
  ## early stopping once the validation AUC doesn't improve by at least 0.01% for 5 consecutive scoring events
  stopping_rounds = 5, stopping_tolerance = 1e-4, stopping_metric = "logloss", 
  
  ## score every 10 trees to make early stopping reproducible (it depends on the scoring interval)
  score_tree_interval = 10,                                                
  
  ## base random number generator seed for each model (automatically gets incremented internally for each model)
  seed = 1234                                                             
)

## Sort the grid models by logloss
sortedGrid <- h2o.getGrid("final_grid", sort_by = "logloss", decreasing = TRUE)    
sortedGrid
```











